---
title: "Final Project (MSC)"
format: html
editor: visual
---

## Final Project: University Dropout

```{r}
#| warning: false

library(tidyverse)
library(tidymodels)
library(janitor)
library(skimr)
library(GGally)
library(broom)
library(corrplot)
library(forcats)
library(scales)
library(purrr)
library(MASS)
library(rstatix)
library(randomForest)
library(caret)
library(pROC)


df <- read.csv("../data/dropout_data.csv", sep = ";")

df <- clean_names(df)

cat_vars <- c(
  "application_mode",
  "course",
  "daytime_evening_attendance",
  "mother_s_qualification",
  "father_s_qualification",
  "mother_s_occupation",
  "father_s_occupation",
  "displaced",
  "debtor",
  "tuition_fees_up_to_date",
  "gender",
  "scholarship_holder"
)

num_vars <- c(
  "application_order",
  "previous_qualification_grade",
  "admission_grade",
  "age_at_enrollment",
  "curricular_units_1st_sem_enrolled",
  "curricular_units_1st_sem_evaluations",
  "curricular_units_1st_sem_approved",
  "curricular_units_1st_sem_grade",
  "curricular_units_2nd_sem_enrolled",
  "curricular_units_2nd_sem_evaluations",
  "curricular_units_2nd_sem_approved",
  "curricular_units_2nd_sem_grade",
  "unemployment_rate",
  "inflation_rate",
  "gdp"
)

outcome_colors <- c(
  "Dropout" = "#D55E00",
  "Enrolled" = "#0072B2",
  "Graduate" = "#009E73"
)


dropout_colors <- c(
  "Dropout" = "#D55E00",
  "No dropout" = "#0072B2"
)

theme_report <- theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )


theme_set(theme_report)
```

Preprocesement of the data, check for missing values.

```{r}
#| warning: false

df <- df %>%
  mutate(
    across(all_of(cat_vars), as.factor),
    target = factor(target, levels = c("Dropout", "Enrolled", "Graduate")),
    dropout_bin = if_else(target == "Dropout", 0, 1),
    dropout_bin_f = factor(dropout_bin, levels = c(0, 1), labels = c("Dropout", "No dropout"))
  )

dim(df)
names(df)
glimpse(df)
skim(df)

colSums(is.na(df))
sum(is.na(df))
```

## Descriptive Analysis:

```{r}
#| warning: false

table(df$dropout_bin_f)
prop.table(table(df$dropout_bin_f))

table(df$target)
prop.table(table(df$target))

ggplot(df, aes(x = target, fill = target)) +
  geom_bar() +
  scale_fill_manual(values = outcome_colors) +
  labs(
    title = "Distribution of academic outcomes",
    x = "Outcome",
    y = "Count"
  )

ggplot(df, aes(x = dropout_bin_f, fill = dropout_bin_f)) +
  geom_bar() +
  scale_fill_manual(values = c("Dropout" = "#D55E00", "No dropout" = "#0072B2")) +
  labs(
    title = "Distribution of binary dropout outcome",
    x = "Binary outcome",
    y = "Count"
  )
```

**\*\*Falta –\> fer per iteració\*\***

```         
```

```{r}

num_summary <- df %>%
  summarise(
    across(
      all_of(num_vars),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        min = ~min(.x, na.rm = TRUE),
        p25 = ~quantile(.x, 0.25, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        p75 = ~quantile(.x, 0.75, na.rm = TRUE),
        max = ~max(.x, na.rm = TRUE)
      )
    )
  ) %>%
  pivot_longer(
    everything(),
    names_to = c("variable", ".value"),
    names_sep = "_(?=[^_]+$)"
  )

num_summary

cat_summary <- df %>%
  dplyr::select(tidyselect::all_of(cat_vars)) %>%
  tidyr::pivot_longer(
    cols = dplyr::everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  dplyr::count(variable, value) %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::arrange(variable, dplyr::desc(n))

cat_summary
```

### Numerical Variables:

```{r}

cor_matrix <- df %>%
  dplyr::select(tidyselect::all_of(num_vars)) %>%
  cor(use = "pairwise.complete.obs")

short_names <- c(
  "App order", "Prev qual grade", "Admission grade", "Age",
   "1st enrolled", "1st eval", "1st approved", "1st grade",
  "2nd enrolled", "2nd eval", "2nd approved", "2nd grade",
  "Unemployment", "Inflation", "GDP"
)

colnames(cor_matrix) <- short_names
rownames(cor_matrix) <- short_names

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.cex = 0.8,
  tl.col = "black",
  number.cex = 0.5
)

num_corr_tests <- map_dfr(
  num_vars,
  \(v) {
    
    test <- cor.test(df[[v]], df$dropout_bin)
    
    tibble(
      variable = v,
      correlation = test$estimate,
      p_value = test$p.value,
      conf_low = test$conf.int[1],
      conf_high = test$conf.int[2]
    )
  }
)

num_corr_tests <- num_corr_tests %>%
  mutate(
    abs_corr = abs(correlation),
    significant = p_value < 0.05
  ) %>%
  arrange(desc(abs_corr))

num_corr_tests

ggplot(num_corr_tests,
       aes(x = reorder(variable, abs_corr), y = correlation,
           fill = significant)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "#0072B2", "FALSE" = "grey70"),
    labels = c("Not Significant", "Significant")
  ) +
  labs(
    title = "Correlation of numerical variables with dropout",
    x = NULL,
    y = "Correlation",
    fill = "Statistical significance"
  ) +
  theme_report

ggplot(num_corr_tests,
       aes(x = reorder(variable, correlation),
           y = correlation)) +
  geom_point(size = 3, color = "#0072B2") +
  geom_errorbar(
    aes(ymin = conf_low, ymax = conf_high),
    width = 0.2
  ) +
  coord_flip() +
  labs(
    title = "Correlation of numerical variables with dropout",
    x = NULL,
    y = "Correlation (95% CI)"
  ) +
  theme_report

key_num_vars <- c(
  "curricular_units_1st_sem_approved",
  "curricular_units_1st_sem_grade",
  "curricular_units_2nd_sem_approved",
  "curricular_units_2nd_sem_grade",
  "age_at_enrollment",
  "curricular_units_2nd_sem_evaluations",
  "curricular_units_2nd_sem_enrolled",
  "admission_grade",
  "previous_qualification_grade"
)

num_compare_tbl <- df %>%
  dplyr::group_by(dropout_bin_f) %>%
  dplyr::summarise(
    dplyr::across(
      tidyselect::all_of(key_num_vars),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE)
      )
    ),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = -dropout_bin_f,
    names_to = c("variable", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  dplyr::arrange(variable, dropout_bin_f)

num_compare_tbl

```

The strongest associations with dropout correspond to academic performance variables, particularly the number of approved units and grades in both semesters. These variables show correlations above 0.45 and are highly statistically significant (p \< 0.001).

Age at enrollment is negatively associated with persistence, indicating that older students are more likely to drop out.

In contrast, macroeconomic indicators such as unemployment rate, inflation rate and GDP exhibit negligible correlations and are not statistically significant.

### Binary Categorical Variables:

```{r}

df_plot <- df %>%
  mutate(
    dropout_bin = factor(dropout_bin,
                         levels = c(0, 1),
                         labels = c("Dropout", "No dropout")),

    attendance = factor(daytime_evening_attendance,
                        levels = c(0, 1),
                        labels = c("Day", "Evening")),

    displaced = factor(displaced,
                       levels = c(0, 1),
                       labels = c("No", "Yes")),

    debtor = factor(debtor,
                    levels = c(0, 1),
                    labels = c("No", "Yes")),

    fees = factor(tuition_fees_up_to_date,
                     levels = c(0, 1),
                     labels = c("Late", "Paid")),

    gender = factor(gender,
                    levels = c(0, 1),
                    labels = c("Female", "Male")),

    scholarship = factor(scholarship_holder,
                         levels = c(0, 1),
                         labels = c("No", "Yes"))
  )


labels = c("Dropout", "No dropout")

cat_vars_small <- c(
  "attendance",
  "displaced",
  "debtor",
  "fees",
  "gender",
  "scholarship"
)


GGally::ggbivariate( 
  df_plot %>% 
    dplyr::select(dropout_bin, tidyselect::all_of(cat_vars_small)),
  outcome = "dropout_bin" ) +
  scale_fill_manual(values = dropout_colors) +
  theme( legend.position = "none" )

binary_tests <- map_dfr(
  cat_vars_small,
  \(v) {
    
    tab <- table(df_plot[[v]], df_plot$dropout_bin)
    
    test <- chisq.test(tab)
    
    tibble(
      variable = v,
      p_value = test$p.value
    )
  }
)

binary_tests
```

Among the binary predictors, financial variables such as tuition payment status, debt and scholarship availability show the strongest associations with dropout.

These results suggest that economic constraints play a key role in student persistence. Gender and attendance schedule also display moderate associations, while the effect of displacement status appears weaker. These variables will therefore be considered in the subsequent logistic regression modelling stage.

### Other Categorical Variables:

```{r}
#| warning: false

df <- df %>%
  mutate(
    dropout_bin_f = factor(dropout_bin,
                           levels = c(0, 1),
                           labels = c("Dropout", "No dropout"))
  )

cat_vars_large <- c(
  "application_mode",
  "course",
  "mother_s_qualification",
  "father_s_qualification",
  "mother_s_occupation",
  "father_s_occupation"
)

cat_tests <- map_dfr(
  cat_vars_large,
  \(v){
    
    tab <- table(df[[v]], df$dropout_bin)
    
    chi <- chisq.test(tab)
    
    tibble(
      variable = v,
      p_value = chi$p.value,
      cramers_v = rstatix::cramer_v(tab)
    )
  }
)

cat_tests %>%
  arrange(desc(cramers_v))

plot_cat_dropout <- function(data, var, top_n = NULL, min_n = NULL, title = NULL) {
  
  df_plot <- data %>%
    dplyr::select(all_of(var), dropout_bin_f) %>%
    mutate(value = as.factor(.data[[var]]))
  
  if (!is.null(top_n)) {
    top_levels <- df_plot %>%
      count(value, sort = TRUE) %>%
      slice_head(n = top_n) %>%
      pull(value)
  
    df_plot <- df_plot %>%
      filter(value %in% top_levels)
  }
  
  if (!is.null(min_n)) {
    valid_levels <- df_plot %>%
      count(value) %>%
      filter(n >= min_n) %>%
      pull(value)
  
    df_plot <- df_plot %>%
      filter(value %in% valid_levels)
  }
  
  df_plot <- df_plot %>%
    group_by(value) %>%
    mutate(drop_rate = mean(
      dropout_bin_f == "Dropout"),
      n = n()) %>%
    ungroup() %>%
    mutate(value = fct_reorder(value, drop_rate))
  
  ggplot(df_plot, aes(x = value, fill = dropout_bin_f)) +
    geom_bar(position = "fill") +
    coord_flip() +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_fill_manual(values = c("Dropout" = "#D55E00", "No dropout" = "#0072B2")) +
    labs(
      title = title,
      x = NULL,
      y = "Proportion"
    ) +
    theme_report +
    theme(legend.position = "bottom")
}

plot_cat_dropout(
  df,
  var = "application_mode",
  top_n = 8,
  title = "Dropout by application mode"
)

plot_cat_dropout(
  df,
  var = "course",
  top_n = 8,
  title = "Dropout by course"
)

plot_cat_dropout(
  df,
  var = "mother_s_qualification",
  min_n = 100,
  title = "Dropout by mother's qualification"
)

plot_cat_dropout(
  df,
  var = "father_s_qualification",
  min_n = 100,
  title = "Dropout by father's qualification"
)

plot_cat_dropout(
  df,
  var = "mother_s_occupation",
  min_n = 100,
  title = "Dropout by mother's occupation"
)

plot_cat_dropout(
  df,
  var = "father_s_occupation",
  min_n = 100,
  title = "Dropout by father's occupation"
)

```

Several categorical predictors exhibit noticeable differences in dropout rates across categories. In particular, application mode and course show substantial heterogeneity, suggesting that both the pathway into higher education and the chosen academic program may influence student persistence. Parental education levels also display moderate associations with dropout, consistent with the role of socio-economic background in educational outcomes. In contrast, parental occupation appears to have a weaker relationship with dropout. These categorical predictors will therefore be included in the subsequent modelling stage to assess their joint effects alongside academic and financial variables.

## Linear models:

(to be discussed)

```{r}

model_1 <- glm(
  dropout_bin ~
    curricular_units_1st_sem_approved +
    curricular_units_1st_sem_grade +
    curricular_units_2nd_sem_approved +
    curricular_units_2nd_sem_grade +
    age_at_enrollment,
  data = df,
  family = binomial
)
```

```{r}

df[cat_vars] <- lapply(df[cat_vars], factor)

all_vars <- c(cat_vars, num_vars)

formula_full <- as.formula(
  paste("dropout_bin ~", paste(all_vars, collapse = " + "))
)

full_model <- glm(
  formula_full,
  data = df,
  family = binomial
)

best_model <- stepAIC(full_model, direction = "both", trace = FALSE)

summary(best_model)
tidy(best_model, exponentiate = TRUE, conf.int = TRUE)
AIC(best_model)
```

## Random Forest:

(to be discussed)

```{r}
# Train/Test split
set.seed(123)

train_index <- createDataPartition(df$dropout_bin_f, p = 0.7, list = FALSE)

train_data <- df[train_index, ]
test_data  <- df[-train_index, ]


# Random Forest model
rf_model <- randomForest(
  dropout_bin_f ~ . - dropout_bin - target,
  data = train_data,
  ntree = 500,
  importance = TRUE
)

# Predictions
rf_predictions <- predict(rf_model, test_data)


# Confusion Matrix
rf_confusion <- confusionMatrix(rf_predictions, test_data$dropout_bin_f)

rf_confusion


# Accuracy
rf_accuracy <- rf_confusion$overall["Accuracy"]
rf_accuracy


# Variable importance
importance_df <- importance(rf_model)

importance_df <- data.frame(
  variable = rownames(importance_df),
  importance = importance_df[,1]
)

importance_df <- importance_df %>%
  arrange(desc(importance)) %>%
  slice(1:15)

ggplot(importance_df,
       aes(x = reorder(variable, importance),
           y = importance)) +
  geom_col(fill = "#0072B2") +
  coord_flip() +
  labs(
    title = "Top variables influencing dropout prediction (Random Forest)",
    x = "",
    y = "Importance"
  ) +
  theme_report


# Logistic comparison
logit_predictions <- predict(best_model, test_data, type = "response")

logit_class <- factor(
  ifelse(logit_predictions > 0.5, "No dropout", "Dropout"),
  levels = c("Dropout","No dropout")
)

confusionMatrix(logit_class, test_data$dropout_bin_f)

```
